Attention Towards Distance Education Tools During COVID-19 Pandemic: Evidence from Google Trends

### 2021-12-16 Paper_online_Learning.Rmd
update online learning

---
title: "Attention Towards Distance Education Tools During COVID-19 Pandemic: Evidence from Google Trends"
author:
- familyname: Talagala 
  othernames: Priyanga Dilini
  address: Department of Compuational Mathematics, University of Moratuwa, Sri Lanka
  email: priyangad@uom.lk
  correspondingauthor: true
- familyname: Talagala
  othernames: Thiyanga
  address: Department of Statistics, University of Sri Jayewardenepura, Sri Lanka
  email: tstalagala@gmail.com
abstract: "A brief summary of our work"
keywords: "rmarkdown, templates"
wpnumber: 1
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  templatehub::workingpaper:
    fig_caption: yes
    fig_height: 10
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

>>>>

---
title: "Attention Towards Distance Education Tools During COVID-19 Pandemic: Evidence from Google Trends"
author:
- familyname: Talagala 
  othernames: Priyanga Dilini
  address: Department of Compuational Mathematics, University of Moratuwa, Sri Lanka
  email: priyangad@uom.lk
  correspondingauthor: true
- familyname: Talagala
  othernames: Thiyanga S.
  address: Department of Statistics, University of Sri Jayewardenepura, Sri Lanka
  email: tstalagala@gmail.com
abstract: "A brief summary of our work"
keywords: "rmarkdown, templates"
wpnumber: 1
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  templatehub::workingpaper:
    fig_caption: yes
    fig_height: 10
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

### 2021-12-16 Paper_online_Learning.tex
update online learning

\author{Priyanga Dilini~Talagala, Thiyanga~Talagala}
\addresses{\textbf{Priyanga Dilini Talagala}\newline
Department of Compuational Mathematics, University of Moratuwa, Sri Lanka
\newline{Email: \href{mailto:priyangad@uom.lk}{\nolinkurl{priyangad@uom.lk}}}\newline Corresponding author\\[1cm]
\textbf{Thiyanga Talagala}\newline
Department of Statistics, University of Sri Jayewardenepura, Sri Lanka
\newline{Email: \href{mailto:tstalagala@gmail.com}{\nolinkurl{tstalagala@gmail.com}}}\\[1cm]
}

@@

The ongoing pandemic of COVID-19 has now become the most important priority of governments and media of many countries all around the world. Due to the alarming levels of spread and severity, the World Health Organization (WHO) declared the outbreak a Public Health Emergency of International (PHEIC) Concern on 30 January 2020, and then a pandemic on 11 March 2020 \autocite{world2020timeline}. As of 29 September 2021, more than 232.7 million cases of COVID-19 have been reported in over 192 countries and territories, resulting in more than 4.7 million deaths \autocite{dong2020interactive}. In response to the pandemic, health authorities worldwide have taken many steps including vaccination development and deployment to minimize the spread of the virus. In addition, authorities worldwide have also taken many non-pharmaceutical interventions and preventive measures such as travel restrictions, lock-downs, workplace hazard controls, school/university closures, facility closures, reduction of mass gatherings to reduce the spread of the virus \autocite{chang2020modelling}.

The first confirmed case of Covid-19 was reported from Sri Lanka on 27 January 2020 with a Chinese national who came to Sri Lanka as a tourist. The first locally transmitted COVID-19 case in Sri Lanka was then reported on 11 March 2020 followed by an increasing trend of the COVID-19 confirmed cases. In response to this increasing trend of COVID 19 cases, lockdowns and curfews were imposed several times across the country with the aim of reducing the spread of the virus. As of 29 September 2021, 515524 cases of COVID-19 have been reported in Sri Lanka, resulting with 12786 deaths \autocite{HealthSL2020,dong2020interactive}.

@@

The ongoing COVID-19 pandemic has brought the world to a standstill. Many sectors have been affected ever since the outbreak of COVID 19, worldwide. Among these many sectors education is one of the most affected sectors with a near-total closures of schools, colleges and universities, all around the world \autocite{daniel2020education}. Figures \ref{fig:covidImpactWorld} shows the evolution of global school closures and reopening since mid February 2020. With the start of the pandemic a near total closure of schools were observed all around the world. In line with that, the Sri Lankan government also ordered to close schools from 12 March 2020 and its planned reopening of schools was delayed several times due to unexpected Covid-19-related circumstances \autocite{wikieducation}. Over time, Africa and Oceania demonstrated better recovery in comparison to other regions with their increasing number of fully open schools (Figures \ref{fig:covidImpactWorld}).

Eventhough distance education has a long history that goes back to almost two centuries \autocite{spector2014handbook} with significant modifications,
alterations, and addition in the process of delivery and communication \autocite{moore2011learning}, COVID 19 certainly made a new era of distance education while encouraging various stakeholders of education to take the concept of distance education seriously. Sudden unexpected movement from classrooms to home schooling at large scale made students, teachers and parents vulnerable, leading to millions of education related internet searches being performed during the period of COVID 19 pandemic (Figures \ref{fig:distanceLearningWorldAnalysis}). Surprisingly, the search spikes of distance education related searches coincide with increasing Covid-19 counts and related internet searches, eventhough distance education has a long history in contrast to Covid-19 pandemic.

@@

\caption{Google trends footprint analysis}\label{fig:distanceLearningWorldAnalysis}
\end{figure}

The speed of these closures and the rapid move to online delivery of education allowed little time for planning or reflection on potential risks that could happen to its various stakeholders. The parents with limited education and resources; and students who tend to have fewer educational opportunities beyond schools and universities severely affected from this sudden unexpected movement. Further, the impact was not limited to students' learning but also to other aspects of their lives such as student debt, digital exclusion, lack of technology, lack of access to internet-enabled device or a stable internet connection, long-term educational disengagement, poor nutrition and food insecurity, increased psychological challenges, childcare problems, exploitation, school dropouts and lack of disability services. Teachers on the other hand were not fully aware of their obligations and how to maintain connections with students to support thier learning. Lack of preparation, lack of tools and techniques for distance education, lack of awareness of the availability of the existing tools and techniques and their effectiveness, lack of interaction and communication with the students, lack of awareness of students' ongoing problems were some of the major issues they encountered during COVID 19 pandemic \autocite{drane2020impact,daniel2020education,unescoadverse2020}. How to preparing course materials, How to interact with students, how to assess and evaluate student work, how to provide feedback are some major concerns they had when preparing for their first online teaching experience. Limited time, and lack of knowledge about the best online tools and techniques available were some of the major challenges teachers encountered during the online course preparation. This also put an extra burden on parents, specially with limited education and resources, as they were expected to facilitate the required learning environment at home.

In response to this grey situation and urgent requirement of massive transformation from physical classroom to virtual learning environment, UNESCO took immediate and timely action by publishing a list of distance learning solutions that can be used to facilitate student learning during the period of school closure. there are many different distance learning solutions. The design of different types of distance learning solutions can depend on the learning objective, target audience, access, and type of content \autocite{moore2011learning}. However, getting familiar with all these available distance solutions is not feasible or practical due to limited time availability and need of urgent responses.

The popularity of a product greatly influences consumer purchasing decisions can indicate its prominence in the market, its usability and its impact \autocite{ahn2006utilizing}. Further, according to \textcite{willis2020using}, developers also tend to improve the their products and services in response to increasing online search. Therefore, the Internet represents a great opportunity to learn about public attention towards a product or service. It's also a way to narrow down thier search space and thereby identify suitable tools and techniques to create an appropriate learning environment.

Google Trends is an open source web analytics tool that allows users to interact with Internet search data, which can provide deep insights into population behavior \autocite{nuti2014use}. In this paper we test whether the Google Analytics search index series can be used as a proxy of the popularity and the public interest in distance education solutions. In line with that, this paper makes three fundamental contributions to distance education by exploring three main questions: (1) What is the impact of COVID 19 pandemic on education? (2) What solutions are in place to meet the need of distance education in terms of coverage, up-take, and learning during COVID 19 pandemic? (3) Which solutions have a wide attention and public interest both at the national and global level?
We primarily analyze quantitative digital footprint data on the Internet from December 2019 to August 2021.The google trend footprint will allow the teachers to narrow down thier search space and select prominent distance education solutions for their teaching purposes.

@@

Notably, GT, together with social media, could be use- fully exploited by global and local health agencies, as well as by public health practitioners, as a tool to identify disease-related information needs, social and behavioural barriers to infection control, and misinformation, as well as to better understand sentiment and risk perception as- sociated with health-threatening events such as an Ebola outbreak {[}5, 37{]}. Indeed, analyses of the temporal trend of GT-based query volumes together with their geographic distribution and main topics of searches can furnish a quantifiable and valuable measure of public attention and information needs about a particular communicable disease \autocite{alicino2015assessing}.

>>>>

\author{Priyanga Dilini~Talagala, Thiyanga S.~Talagala}
\addresses{\textbf{Priyanga Dilini Talagala}\newline
Department of Compuational Mathematics, University of Moratuwa, Sri Lanka
\newline{Email: \href{mailto:priyangad@uom.lk}{\nolinkurl{priyangad@uom.lk}}}\newline Corresponding author\\[1cm]
\textbf{Thiyanga S. Talagala}\newline
Department of Statistics, University of Sri Jayewardenepura, Sri Lanka
\newline{Email: \href{mailto:tstalagala@gmail.com}{\nolinkurl{tstalagala@gmail.com}}}\\[1cm]
}

@@

The ongoing pandemic of COVID-19 is still one of the most important priorities of governments and media of many countries all around the world. Due to the alarming levels of spread and severity, the World Health Organization (WHO) declared the outbreak a Public Health Emergency of International Concern (PHEIC) on 30 January 2020, and then a pandemic on 11 March 2020 \autocite{world2020timeline}. As of 29 September 2021, more than 232.7 million cases of COVID-19 have been reported in over 192 countries and territories, resulting in more than 4.7 million deaths \autocite{dong2020interactive}. In response to the pandemic, health authorities worldwide have taken many steps including vaccination development and deployment to minimize the spread of the virus. In addition, authorities worldwide have also taken many non-pharmaceutical interventions and preventive measures such as travel restrictions, lock-downs, workplace hazard controls, school/university closures, facility closures, reduction of mass gatherings to reduce the spread of the virus \autocite{chang2020modelling}.

The first confirmed case of Covid-19 was reported from Sri Lanka on 27 January 2020 with a person who came to Sri Lanka as a tourist. The first locally transmitted COVID-19 case in Sri Lanka was then reported on 11 March 2020 followed by an increasing trend of the COVID-19 confirmed cases. In response to this increasing trend of COVID 19 cases, lockdowns and curfews were imposed several times across the country with the aim of reducing the spread of the virus. As of 29 September 2021, 515524 cases of COVID-19 have been reported in Sri Lanka, resulting with 12786 deaths \autocite{HealthSL2020,dong2020interactive}.

@@

The ongoing COVID-19 pandemic has brought the world to a standstill. Many sectors have been affected ever since the outbreak of COVID 19, worldwide. Among these many sectors education is one of the most affected sectors with a near-total closures of schools, colleges and universities, all around the world \autocite{daniel2020education}. Figures \ref{fig:covidImpactWorld} shows the evolution of global school closures and reopening since mid February 2020. With the start of the pandemic a near total closure of schools were observed all around the world. In line with that, the Sri Lankan government also ordered to close schools from 12 March 2020 and its planned reopening of schools was delayed several times due to unexpected Covid-19-related circumstances \autocite{wikieducation}. Over time, Africa and Oceania demonstrated better recovery in comparison to other regions with their increasing number of fully open schools (Figures \ref{fig:covidImpactContinent}).

@@

\caption{Visualization of data from Goolge search trends. (a) COVID-19 cases worldwide (b) Google search trends of COVID-19 related terms (c) Google search trends of distance education related terms.}\label{fig:distanceLearningWorldAnalysis}
\end{figure}

Eventhough distance education has a long history that goes back to almost two centuries with significant modifications,
alterations, and addition in the process of delivery and communication \autocite{moore2011learning,spector2014handbook}, COVID 19 certainly made a new era of distance education while encouraging various stakeholders of education to take the concept of distance education seriously. Sudden unexpected movement from classrooms to home schooling at large scale made students, teachers and parents vulnerable, leading to millions of education related internet searches being performed during the period of COVID 19 pandemic (Figures \ref{fig:distanceLearningWorldAnalysis}). Surprisingly, the search spikes of distance education related searches (Figure \ref{fig:distanceLearningWorldAnalysis} (c)) coincide with increasing Covid-19 counts (Figure \ref{fig:distanceLearningWorldAnalysis} (a)) and related internet searches (Figure \ref{fig:distanceLearningWorldAnalysis} (b)), eventhough distance education has a long history in contrast to Covid-19 pandemic.

The speed of the physical classroom closures and the rapid move to online delivery of education allowed little time for planning or reflection on potential risks that could happen to its various stakeholders: teachers, students and parents. Teachers were not fully aware of their obligations and how to maintain connections with students to support thier learning. Lack of preparation, lack of tools and techniques for distance education, lack of awareness of the availability of the existing tools and techniques and their effectiveness, lack of interaction and communication with the students, lack of awareness of students' ongoing problems were some of the major issues they encountered during COVID 19 pandemic. Students on the other hand who tend to have fewer educational opportunities beyond schools and universities severely affected from this sudden unexpected movement. Further, the impact was not limited to students' learning but also to other aspects of their lives such as student debt, digital exclusion, lack of technology, lack of access to internet-enabled device or a stable internet connection, long-term educational disengagement, poor nutrition and food insecurity, increased psychological challenges, childcare problems, exploitation, school dropouts and lack of disability services. \autocite{drane2020impact,daniel2020education,unescoadverse2020}. This also put an extra burden on parents, specially with limited education and resources, as they were expected to facilitate the required learning environment at home. Further, every single step including planning, developing new tools and techniques, conducting awareness programmes about distance education, shifting towards distance education, happened through online, due to the unexpected massive global shutdown. This situation left internet as the only medium to support all these educational processes.

In response to this gray situation and urgent requirement of massive transformation from physical classroom to virtual learning environment, UNESCO took immediate and timely action by publishing a long list of distance learning solutions that can be used to facilitate student distance learning during the period of school closure. There are many different distance learning solutions. The design of different types of distance learning solutions can depend on the learning objective, target audience, access, and type of content \autocite{moore2011learning}. However, getting familiar with all these available distance solutions is not feasible or practical due to limited time availability and need of urgent responses.

According to \textcite{ahn2006utilizing}, the popularity of a product greatly influences consumer purchasing decisions. It can also indicate the prominence of the product in the market, its usability and its impact. Further, according to \textcite{willis2020using}, developers also tend to improve their products and services in response to increasing online search. Therefore, the internet represents a great opportunity to learn about public attention towards a product or service. It is also a way to narrow down the search space and thereby identify most suitable product and services that meets customer needs.

Google Trends is an open source web analytics tool that allows users to interact with Internet search data, which can provide deep insights into population behavior \autocite{nuti2014use}. In this study we test whether the Google Analytics search index series can be used as a proxy of the popularity and the public interest in different distance education solutions. In line with that, this paper makes three fundamental contributions to distance education by exploring three main questions: (1) What is the impact of COVID 19 pandemic on education? (2) What solutions are in place to meet the need of distance education in terms during COVID 19 pandemic? (3) Which distance learning solutions have a wide attention and public interest both at the national and global level?
We primarily analyze quantitative digital footprint data on the internet from December 2019 to August 2021. The resulted google trend footprint provides a fast first step to identify the most popular distance education tools available for different education purposes. It also allows the teachers to narrow down the search space and deepen their exploration on prominent distance education solutions to support their online teaching.

Quality education is essential to sustainable development and it is a key United Nations (UN) sustainable development goal. Having the right tool is imperative to successful completion of a task at hand. The findings of this study provides an initial guidance to select a right tool. Therefore, the findings of this study directly contribute to UN's sustainable development goals of quality education.

@@

Notably, GT, together with social media, could be use- fully exploited by global and local health agencies, as well as by public health practitioners, as a tool to identify disease-related information needs, social and behavioural barriers to infection control, and misinformation, as well as to better understand sentiment and risk perception associated with health-threatening events such as an Ebola outbreak {[}5, 37{]}. Indeed, analyses of the temporal trend of GT-based query volumes together with their geographic distribution and main topics of searches can furnish a quantifiable and valuable measure of public attention and information needs about a particular communicable disease \autocite{alicino2015assessing}.

### 2021-12-16 main.Rmd
update online learning

The ongoing pandemic of COVID-19 has now become the most important priority of governments and  media of many countries all around the world.  Due to the alarming levels of spread and severity, the World Health Organization (WHO) declared the outbreak a Public Health Emergency of International (PHEIC) Concern on 30 January 2020, and then a pandemic on 11 March 2020 [@world2020timeline]. As of 29 September 2021, more than 232.7 million cases of COVID-19 have been reported in over 192 countries and territories, resulting in more than 4.7 million deaths [@dong2020interactive].  In response to the pandemic, health authorities worldwide have taken many steps including vaccination development and deployment  to minimize the spread of the virus. In addition, authorities worldwide have also taken many non-pharmaceutical interventions and preventive measures such as  travel restrictions, lock-downs, workplace hazard controls, school/university closures, facility closures, reduction of mass gatherings to reduce the spread of the virus [@chang2020modelling].

The first confirmed case of Covid-19 was reported from Sri Lanka on 27 January 2020 with  a Chinese national who came to Sri Lanka as a tourist.  The first locally transmitted COVID-19 case in Sri Lanka was then reported on  11 March 2020 followed by an increasing trend of the COVID-19 confirmed cases. In response to this increasing trend of COVID 19 cases, lockdowns and curfews were imposed several times across the country with the aim of reducing the spread of the virus. As of 29 September 2021, 515524 cases of COVID-19 have been reported in Sri Lanka, resulting with 12786 deaths [@HealthSL2020; @dong2020interactive].

@@

The ongoing COVID-19 pandemic has brought the world to a standstill.  Many sectors have been affected ever since the outbreak of COVID 19, worldwide. Among these many sectors education is one of the most affected sectors with a near-total closures of schools, colleges and universities, all around the world [@daniel2020education]. Figures \ref{fig:covidImpactWorld} shows the evolution of global school closures and reopening since mid February 2020. With the start of the pandemic a near total closure of schools were observed all around the world. In line with that, the Sri Lankan  government also ordered to close schools from 12 March 2020 and  its planned reopening of schools was delayed several times due to unexpected Covid-19-related circumstances [@wikieducation]. Over time, Africa and Oceania demonstrated better recovery in comparison to other regions with their increasing number of fully open schools (Figures \ref{fig:covidImpactWorld}).

Eventhough distance education has a long history that goes back to almost two centuries [@spector2014handbook] with significant  modifications,
alterations, and addition in the process of delivery and communication  [@moore2011learning],  COVID 19 certainly made a new era of distance education while encouraging various stakeholders of education to take the concept of distance education seriously.  Sudden unexpected movement from classrooms to home schooling at large scale made  students, teachers and parents vulnerable, leading to millions of education related internet searches being performed during the period of COVID 19 pandemic (Figures \ref{fig:distanceLearningWorldAnalysis}). Surprisingly, the search spikes of distance education related searches coincide with increasing Covid-19 counts and related internet searches, eventhough distance education has a long history in contrast to Covid-19 pandemic.

```{r distanceLearningWorldAnalysis, fig.cap= "Google trends footprint analysis", fig.height= 8}
```

The speed of these closures and the rapid move to  online delivery of education allowed little time for planning or reflection on potential risks that could happen to its various stakeholders.   The parents  with limited education and  resources;  and  students  who tend to have fewer educational opportunities beyond schools and universities severely affected from this sudden unexpected movement. Further, the impact was not limited to students' learning but also to other aspects of their lives such as  student debt, digital exclusion,  lack of technology, lack of access to internet-enabled device or a stable internet connection, long-term educational disengagement,  poor nutrition and food insecurity, increased psychological challenges, childcare problems, exploitation, school dropouts  and lack of disability services. Teachers on the other hand were not fully aware of their obligations and how to maintain connections with students to support thier learning. Lack of preparation, lack of tools and techniques for distance education, lack of  awareness of the availability of the existing  tools and techniques and their effectiveness, lack of interaction and  communication with the students, lack of awareness of students' ongoing problems  were some of the major issues they encountered during COVID 19 pandemic [@drane2020impact; @daniel2020education; @unescoadverse2020]. How to preparing course materials, How to interact with students, how to  assess and evaluate student work, how to provide feedback are some major concerns they had when preparing for their first  online  teaching experience. Limited time, and lack of knowledge about the best online tools and techniques available were some of the major challenges teachers encountered during the online course preparation. This also put an extra burden on parents, specially with limited education and resources,  as they were  expected to facilitate the required learning environment at home.

In response to this grey situation and urgent requirement of massive transformation from physical classroom to virtual learning environment,  UNESCO took immediate and timely action by publishing a list of distance learning solutions that can be used to facilitate student learning during the period of school closure. there are many different distance learning solutions.  The design of different types of distance learning solutions can depend on the learning objective, target audience, access, and type of content [@moore2011learning]. However, getting familiar with all these available distance solutions  is  not feasible or practical due to limited time availability and need of urgent responses.

The popularity of a product greatly influences consumer purchasing decisions can indicate its prominence in the market, its usability and its impact  [@ahn2006utilizing]. Further, according to @willis2020using, developers also tend to improve the their products and services in response to increasing online search. Therefore, the Internet represents a great opportunity to learn about public attention towards a product or service.  It's also a way to narrow down thier search space and  thereby identify suitable tools and techniques to create an appropriate learning environment.

Google Trends is an open source web analytics tool that allows users to interact with Internet search data, which can provide deep insights into population behavior [@nuti2014use]. In this paper we test whether the Google Analytics search index series can be used as a proxy of the popularity and the public interest in distance education solutions. In line with that, this paper makes three fundamental contributions to distance education by exploring three main questions: (1) What is the impact of  COVID 19 pandemic on education? (2) What solutions are in place to meet the need of distance education in terms of coverage, up-take, and learning during COVID 19 pandemic? (3) Which solutions have a wide attention and public interest both at the national and global level? 
We primarily analyze quantitative digital footprint data on the Internet from December 2019 to August 2021.The google trend footprint will allow the teachers to narrow down thier search space and select  prominent distance education solutions for their teaching purposes.

@@

Notably, GT, together with social media, could be use- fully exploited by global and local health agencies, as well as by public health practitioners, as a tool to identify disease-related information needs, social and behavioural barriers to infection control, and misinformation, as well as to better understand sentiment and risk perception as- sociated with health-threatening events such as an Ebola outbreak [5, 37]. Indeed, analyses of the temporal trend of GT-based query volumes together with their geographic distribution and main topics of searches can furnish a quantifiable and valuable measure of public attention and information needs about a particular communicable disease [@alicino2015assessing].

>>>>

The ongoing pandemic of COVID-19 is still one of the most important priorities of governments and  media of many countries all around the world.  Due to the alarming levels of spread and severity, the World Health Organization (WHO) declared the outbreak a Public Health Emergency of International Concern (PHEIC) on 30 January 2020, and then a pandemic on 11 March 2020 [@world2020timeline]. As of 29 September 2021, more than 232.7 million cases of COVID-19 have been reported in over 192 countries and territories, resulting in more than 4.7 million deaths [@dong2020interactive].  In response to the pandemic, health authorities worldwide have taken many steps including vaccination development and deployment  to minimize the spread of the virus. In addition, authorities worldwide have also taken many non-pharmaceutical interventions and preventive measures such as  travel restrictions, lock-downs, workplace hazard controls, school/university closures, facility closures, reduction of mass gatherings to reduce the spread of the virus [@chang2020modelling].

The first confirmed case of Covid-19 was reported from Sri Lanka on 27 January 2020 with  a person who came to Sri Lanka as a tourist.  The first locally transmitted COVID-19 case in Sri Lanka was then reported on  11 March 2020 followed by an increasing trend of the COVID-19 confirmed cases. In response to this increasing trend of COVID 19 cases, lockdowns and curfews were imposed several times across the country with the aim of reducing the spread of the virus. As of 29 September 2021, 515524 cases of COVID-19 have been reported in Sri Lanka, resulting with 12786 deaths [@HealthSL2020; @dong2020interactive].

@@

The ongoing COVID-19 pandemic has brought the world to a standstill.  Many sectors have been affected ever since the outbreak of COVID 19, worldwide. Among these many sectors education is one of the most affected sectors with a near-total closures of schools, colleges and universities, all around the world [@daniel2020education]. Figures \ref{fig:covidImpactWorld} shows the evolution of global school closures and reopening since mid February 2020. With the start of the pandemic a near total closure of schools were observed all around the world. In line with that, the Sri Lankan  government also ordered to close schools from 12 March 2020 and  its planned reopening of schools was delayed several times due to unexpected Covid-19-related circumstances [@wikieducation]. Over time, Africa and Oceania demonstrated better recovery in comparison to other regions with their increasing number of fully open schools (Figures \ref{fig:covidImpactContinent}).

```{r distanceLearningWorldAnalysis, fig.cap= "Visualization of data from Goolge search trends. (a) COVID-19 cases worldwide (b) Google search trends of COVID-19 related terms (c) Google search trends of distance education related terms.", fig.height= 8}
```

Eventhough distance education has a long history that goes back to almost two centuries with significant  modifications,
alterations, and addition in the process of delivery and communication  [@moore2011learning; @spector2014handbook],  COVID 19 certainly made a new era of distance education while encouraging various stakeholders of education to take the concept of distance education seriously.  Sudden unexpected movement from classrooms to home schooling at large scale made  students, teachers and parents vulnerable, leading to millions of education related internet searches being performed during the period of COVID 19 pandemic (Figures \ref{fig:distanceLearningWorldAnalysis}). Surprisingly, the search spikes of distance education related searches (Figure \ref{fig:distanceLearningWorldAnalysis} (c)) coincide with increasing Covid-19 counts (Figure \ref{fig:distanceLearningWorldAnalysis} (a)) and related internet searches (Figure \ref{fig:distanceLearningWorldAnalysis} (b)), eventhough distance education has a long history in contrast to Covid-19 pandemic.

The speed of the physical classroom closures and the rapid move to  online delivery of education allowed little time for planning or reflection on potential risks that could happen to its various stakeholders: teachers, students and parents.   Teachers were not fully aware of their obligations and how to maintain connections with students to support thier learning. Lack of preparation, lack of tools and techniques for distance education, lack of  awareness of the availability of the existing  tools and techniques and their effectiveness, lack of interaction and  communication with the students, lack of awareness of students' ongoing problems  were some of the major issues they encountered during COVID 19 pandemic. Students  on the other hand  who tend to have fewer educational opportunities beyond schools and universities severely affected from this sudden unexpected movement. Further, the impact was not limited to students' learning but also to other aspects of their lives such as  student debt, digital exclusion,  lack of technology, lack of access to internet-enabled device or a stable internet connection, long-term educational disengagement,  poor nutrition and food insecurity, increased psychological challenges, childcare problems, exploitation, school dropouts  and lack of disability services. [@drane2020impact; @daniel2020education; @unescoadverse2020]. This also put an extra burden on parents, specially with limited education and resources,  as they were  expected to facilitate the required learning environment at home. Further, every single step including planning, developing new tools and techniques, conducting awareness programmes about distance education, shifting towards distance education, happened through online, due to the unexpected massive global shutdown. This situation left internet as the only medium to support all these educational processes.

In response to this gray situation and urgent requirement of massive transformation from physical classroom to virtual learning environment,  UNESCO took immediate and timely action by publishing a long list of distance learning solutions that can be used to facilitate student distance learning during the period of school closure. There are many different distance learning solutions.  The design of different types of distance learning solutions can depend on the learning objective, target audience, access, and type of content [@moore2011learning]. However, getting familiar with all these available distance solutions  is  not feasible or practical due to limited time availability and need of urgent responses.

According to @ahn2006utilizing, the popularity of a product greatly influences consumer purchasing decisions.  It can also indicate the prominence of the product in the market, its usability and its impact.  Further, according to @willis2020using, developers also tend to improve their products and services in response to increasing online search. Therefore, the internet represents a great opportunity to learn about public attention towards a product or service.  It is also a way to narrow down the search space and  thereby identify most suitable product and services that meets customer needs.

Google Trends is an open source web analytics tool that allows users to interact with Internet search data, which can provide deep insights into population behavior [@nuti2014use]. In this study we test whether the Google Analytics search index series can be used as a proxy of the popularity and the public interest in different distance education solutions. In line with that, this paper makes three fundamental contributions to distance education by exploring three main questions: (1) What is the impact of  COVID 19 pandemic on education? (2) What solutions are in place to meet the need of distance education in terms during COVID 19 pandemic? (3) Which distance learning solutions have a wide attention and public interest both at the national and global level? 
We primarily analyze quantitative digital footprint data on the internet from December 2019 to August 2021. The resulted google trend footprint provides a fast first step to identify the most popular distance education tools available for different education purposes. It  also allows the teachers to narrow down the search space and  deepen their exploration on  prominent distance education solutions to support their online teaching.

Quality education is essential to sustainable development and it is a key  United Nations (UN) sustainable development goal. Having the right tool is imperative to successful completion of a task at hand. The findings of this study provides an initial guidance to select a right tool. Therefore, the findings of this study directly contribute to UN's sustainable development goals of quality education.

@@

Notably, GT, together with social media, could be use- fully exploited by global and local health agencies, as well as by public health practitioners, as a tool to identify disease-related information needs, social and behavioural barriers to infection control, and misinformation, as well as to better understand sentiment and risk perception associated with health-threatening events such as an Ebola outbreak [5, 37]. Indeed, analyses of the temporal trend of GT-based query volumes together with their geographic distribution and main topics of searches can furnish a quantifiable and valuable measure of public attention and information needs about a particular communicable disease [@alicino2015assessing].
